Asymptotic and Numerical Solution for Electro-Kinetic Flow Around an Ion-Selective Particle

### 2012-09-08 notes/work/ijnam.tex
Work on IJNAM.

\title{Numerical and analytical solution of nonlinear electro-kinetic phenomena}

@@

% dedication
\dedicatory{This paper is dedicated to our authors}

@@

\abstract{A numerical solver for the simulation of ion-exchanging particles migration in an
electrolyte solution is presented. The electrokinetic transport process is
described by a nonlinear partial differential equation system, 
as well as nonlinear boundary conditions.
The nonlinear system is discretized and solved iteratively using Newton's method.
An asymptotic analytical solution is derived for the validation of the solver. 
The numerical results are compared to the asymptotic solutions, and good correspondence is achieved.}

@@

Electrokinetic theory describes the dynamics of charged particles
in ionic fluids \cite{masliyah2005book,kirby2010book}.
When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via    
electric forces, creating a double-layer structure around the
particle (see Figure \ref{fig:EDL}). This structure, called
``Debye layer'', electrically screens the surface charge, thus
creating a potential difference between the particle and the outer
layer of the fluid bulk.
In cases where the layer width is much smaller than the particle's
size, an analytical asymptotic solution for the Debye layer
dynamics can be found \cite{yariv2010asymptotic}.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=0.2\textwidth]
            {figs/ElectricDoubleLayer.eps}
        \caption{Schematic structure of the double layer:
        (1) particle surface, (2) Debye layer, (3) fluid bulk.
        If the particle surface has positive (blue) surface charge,
        it attracts negative (red) ions from the fluid making the
        Debye layer negatively charged (as opposed to the rest of
        the fluid bulk, which is electrically neutral).
        The zeta potential is defined as the voltage drop across (2).}
        \label{fig:EDL}
    \end{center}
\end{figure}
% \begin{tikzpicture}\end{tikzpicture}

The variables of the electrokinetic problem are the electrostatic
potential $\varphi$, fluid velocity $\bv$ and its pressure $p$, and
ionic concentration $c$.
The boundary conditions are determined by the specific
problem under consideration and defined by the particle's
geometry, chemical characteristics, and the fluid dynamics.
The partial differential equations that describe the system dynamics
under an external electric field, are coupled and nonlinear, and
in general they have no analytic solution. 
Moreover, any numerical solver must handle the scale disparity caused by the
Debye layer width being much smaller than the particle's size.

A closed-form linear asymptotic solution has been developed for
spherical ion-exchanging particles and weak electric field in an
axisymmetric setting \cite{yariv2010migration} but it is hard to extend this
analytic solution to more general systems. Once the electric field
becomes stronger, significant nonlinear phenomena are expected.
This interesting regime has not yet been explored.

Recently, it has been conceived that such type of dynamics
can be very useful in transporting and manipulating micro-
and nanoscale objects in many nanotechnology applications,
such as the self-assembly of superstructures, roving sensors, 
drug-delivery systems and useful nanomachinery 
\cite{howse2007self,paxton2004catalytic,pumera2010electrochemically}.

However, since the model for these electrokinetic
phenomena has no closed-form analytic solution, a numerical
solver for this system is of importance.

The goal of this work is to develop and implement an iterative numerical
solver for nonlinear electrokinetic problems, and apply it to the study
of systems that have no closed-form solutions.
This solver can be used to gain insight into the chemical and physical behavior 
in far more general regimes than are currently well-understood, 
and also to possibly lead to further analytical developments, based on new scaling behavior
discovered numerically. We may also target interesting physical
phenomena that have been observed experimentally.
One such phenomenon is nonlinear electrokinetic vortex flow, 
which can useful for microfluidic mixing applications \cite{wang2004mix, ben2002vortex},

The paper is structured as follows: the governing equations of the electrokinetic model
are derived in Section \ref{sec:equations}. The algorithm for the discretization and 
the solution of the system is described in Section \ref{sec:algorithm}.
The analytical asymptotic solution for the non-linear system is derived 
and presented in Section \ref{sec:asymp}.
The numerical results and their comparison to the analytical asymptotic solution 
are presented and discussed in Section \ref{sec:results}.

@@

Spatial coordinates $\br$ are normalized by $a^*$ 
(the characteristic size of the ion-exchanging particle).

Positive and negative ionic concentrations are denoted by $c_+$ and $c_-$, respectively, and
are normalized by the ambient ionic concentration $c^*$. The ionic valences are $\pm z$.

The dimensional ionic fluxes, $\bj_\pm$, are given by the Nernst-Plank equation 
(using the Einstein relation for the electric mobility, $\mu_q^* = D^* / k_B^* T^*$):
\begin{eqnarray}
\bj^*_\pm &=& 
-D^* \bnabla c^*_\pm + \bv^* c^*_\pm \mp \frac{z e^* D^*}{k_B^* T^*} c^*_\pm \bnabla \varphi^*,
\end{eqnarray}
where $D^*$ is the ionic diffusivity.

@@

The dimensional Poisson equation can be written as:
\begin{eqnarray}
(c^*_+ - c^*_-) z F^* &=& -\eps^* \Laplacian \varphi^*,
\end{eqnarray}
and after nondimensionalization, is given by:
\begin{eqnarray}
c_+ - c_- &=& -\frac{\eps^* \varphi^*}{z F^* c^* (a^*)^2} \Laplacian \varphi = 
-\delta^2 \Laplacian \varphi,
\\
\delta^2 &=& \frac{\eps^* \varphi^*}{c^* F^* z (a^*)^2} = 
\pars{\frac{\delta^*}{a^*}}^2.
\end{eqnarray}

@@

\subsection{Micro-scale phenomena}

@@

Due to conservation of ions, the fluxes are divergence-free:
\begin{eqnarray}
\bnabla \cdot \bj_\pm &=& 0.
\end{eqnarray}
These equation may be re-written by using the following transformations:
\begin{eqnarray}
  c &=& \frac{c_+ + c_-}{2},\\
  q &=& \frac{c_+ - c_-}{2}.
\end{eqnarray}
Thus, salt and charge fluxes are defined by:
\begin{eqnarray}
  \bj &=& \frac{\bj_+ + \bj_-}{2} = -\bnabla c - q \bnabla \varphi + \alpha \bv c, \\
  \bi &=& \frac{\bj_+ - \bj_-}{2} = -\bnabla q - c \bnabla \varphi + \alpha \bv q,
\end{eqnarray}
and are divergence-free as well:
\begin{eqnarray}
\bnabla \cdot \bj &=& 0, \\
\bnabla \cdot \bi &=& 0. 
\end{eqnarray}

@@

\subsubsection{Boundary conditions}
We follow the physical setting described in \cite{yariv2010migration}.
We assume that a constant constant electric field of magnitude $\beta$,
is applied to the system, in direction denoted by the unit vector $\ui$.
As a result, the ion-exchanger will equilibrate in 
a non-zero steady-state velocity $\cU \ui$, which is
a function of the applied field, and is not known a priori.

An ion-exchanger fixed reference frame is used.
Far away from the particle, we have a uniform applied electrostatic field, uniform flow,
and ambient ionic concentrations:
\begin{eqnarray}
-\bv &\rightarrow& \cU \ui, \\
-\bnabla \varphi &\rightarrow& \beta\ui, \\
c_\pm &\rightarrow& 1.
\end{eqnarray}

We consider a spherical ion-exchanger. 
At its surface $\mathcal{S}$ (defined at $r=1$ with normal $\bn = \brhat$), 
we assume high conductance of the particle (yielding a uniform electric potential),
zero slip, anion impermeability, cation selectivity 
(with fast Butler-Volmer\cite{bard2000book} kinetics
$\bn \cdot \bj_+ = k (1 - c_+/\gamma)$, where $k \rightarrow \infty$):
\begin{eqnarray}
\varphi & = & \cV, \\
\bv & = & \bzero, \\
\bn \cdot \bj_- & = & 0, \\
c_+ &=& \gamma, \\
\end{eqnarray}
where $\gamma$ is the ratio between the cationic concentration on the surface and the cationic
concentration in the fluid bulk.

@@

\subsubsection  {Bulk-scale equations}
The Nernst-Planck equations above for the fluid bulk (outside
the boundary layer), using bulk variables denoted by uppercase letters, read:
\begin{eqnarray}
  Q & = & 0, \\
  C & = & C_+ = C_-, \\
\bJ &=& -\bnabla C + \alpha \bV C, \\
\bI &=& -C \bnabla \varPhi.
\end{eqnarray}
Because the flow is incompressible ($\bnabla \cdot V = 0$), 
ion conservation results in the following equations:

\begin{enumerate}
\item Salt conservation in macroscale equation:
\begin{eqnarray} \label{eq:salt}
\bnabla \cdot \bJ = \Laplacian C - \alpha \bV \cdot \bnabla C = 0. 
\end{eqnarray}

\item Charge conservation in macroscale equations:
\begin{eqnarray} \label{eq:charge}
\bnabla \cdot \bI = \bnabla \cdot \pars{ C \bnabla \varPhi } = 0.
\end{eqnarray}

\end{enumerate}

@@

\subsubsection{Bulk-scale effective boundary conditions}
By integrating the microscale equations across the Debye layer
(see the full derivation in \cite{yariv2010asymptotic}), 
the effective boundary conditions at the ion-exchanger surface are:
\begin{eqnarray}
0 &=& \varPhi + \log C, \\
0 &=& \bnhat \cdot \bnabla \pars{\varPhi - \log C}, \\
\bV &=& 
\zeta \cdot \bnabla_\mathcal{S} \varPhi 
+ 2\log\pars{1-\tanh^2\frac{\zeta}{4}} \cdot \bnabla_\mathcal{S} \log C,
\end{eqnarray}
where $\zeta = \cV - \varPhi = \log (C / \gamma)$ 
is the zeta potential (the voltage drop across the Debye layer).

\subsection{Steady state}
The stress tensor is composed of Newtonian and Maxwel stresses:
\begin{eqnarray}
\tT &=& \bnabla \bV + (\bnabla \bV)^\dagger - P \tI
+ \bnabla \varPhi \bnabla \varPhi - \frac{1}{2} (\bnabla \varPhi \cdot \bnabla \varPhi) \tI, \\
\bnabla \cdot \tT &=& \bzero.
\end{eqnarray}
The total force acting on the particle is computed by the following surface integral:
\begin{eqnarray}
\bF &=& \oint_\mathcal{S} \tT \cdot \bnhat dA.
\end{eqnarray}
In steady state (constant particle drift velocity $\cU$), 
the total force must be zero: 
\begin{eqnarray}
\bF = \bzero.
\end{eqnarray}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algorithm Description} \label{sec:algorithm}
The numerical solver first requires constructing a discrete 
approximation for the nonlinear PDE system,
for a given electric field $\beta$ and a velocity $\cU$. 
The resulting nonlinear discrete system is
solved numerically using Newton's Method, so the total force on the particle can be computed
numerically. Variation of $\cU$ is used to find the steady-state velocity, 
which corresponds to zero total force.

The steady-state velocity is thus the solution to the nonlinear scalar equation
\begin{eqnarray}
F(\beta, \cU) = 0,
\end{eqnarray}
which we solver by standard secant method.

\subsection{Discretization}
\subsubsection{Spherical Coordinates}
Since the system is axisymmetric, the operators are written in 
spherical coordinates $(r,\theta,\phi)$.
The scalar gradient and the divergence can be written as:
\begin{eqnarray}
\bnabla f &=& \deriv{f}{r}\brhat + \frac{1}{r}\deriv{f}{\theta}\bthetahat, \\
\bnabla \cdot \bF &=& \frac{1}{r^2}\deriv{}{r}\pars{F_r r^2 } + 
               \frac{1}{r \sin\theta}\deriv{}{\theta}\pars{F_\theta \sin\theta}.
\end{eqnarray}
The scalar Laplacian can be written as:
\begin{eqnarray}
\Laplacian f = \bnabla \cdot (\bnabla f)&=& 
 \frac{1}{r^2}\deriv{}{r}\pars{r^2 \deriv{f}{r}} + 
 \frac{1}{r^2 \sin\theta}\deriv{}{\theta}\pars{\sin\theta \deriv{f}{\theta}}.
\end{eqnarray}
The vector gradient can be written as:
\begin{eqnarray}
\bnabla \bF &=& \deriv{F_r}{r} \brhat \brhat + \deriv{F_\theta}{r} \brhat \bthetahat + 
\frac{1}{r}\pars{\deriv{F_r}{\theta} - F_\theta} \bthetahat \brhat + 
\frac{1}{r}\pars{\deriv{F_\theta}{\theta} + F_r} \bthetahat \bthetahat.
\end{eqnarray}
The vector Laplacian can be written as:
\begin{eqnarray}
\bLaplacian \bF &=& 
\left(\Laplacian F_r - \frac{2F_r}{r^2} - 
\frac{2}{r^2 \sin\theta} \deriv{\left(F_\theta \sin\theta \right)}{\theta}\right)\brhat
+ \left(\Laplacian F_\theta - \frac{F_\theta}{r^2 \sin^2\theta} + 
\frac{2}{r^2}\deriv{F_r}{\theta}\right) \bthetahat.
\end{eqnarray}

It is important to note that due to the choice of spherical coordinate system,
the vector Laplacian operator cannot be computed in a decoupled manner as in the Cartesian system,
since the unit vectors $\brhat, \bthetahat, \bphihat$ are functions of the coordinates
in a curvilinear coordinate system.

\subsubsection{Computational Grid}
In order to discretize the differential operators, 
a regular grid of size $n_r \times n_\theta$ is defined, with:
\begin{eqnarray}
(r_i,\theta_j) &\in& [1, \infty) \times [0,\pi], \\ 
r_i &=& (1+\Delta_r)^i, \\
\theta_j &=& \Delta_\theta \cdot j,
\end{eqnarray}
where a logarithmic grid spacing has been chosen for $r$ and a uniform grid for $\theta$.
Note that $r_0 = 1$ and $R_{max} = (1+\Delta_r)^{n_r}$ should be much 
larger than $r_0$. Thus:
\begin{eqnarray}
\Delta_r &=& \pars{R_{max}} ^ \frac{1}{n_r} - 1, \\
\Delta_\theta &=& \frac{\pi}{n_\theta}.
\end{eqnarray}

This grid induces a disjoint subdivision of the domain 
$\Omega = [1, R_{max}] \times [0,\pi]$ into cells:
\begin{eqnarray}
\bigcup_{ij}\Omega_{ij} &=& \Omega.
\end{eqnarray}
A specific cell $\Omega_{ij}$ and its center $(\bar{r}_i, \bar{\theta}_j)$ are defined by:
\begin{eqnarray}
\Omega_{ij} &=& [r_{i-1}, r_{i}] \times [\theta_{j-1}, \theta_{j}], \\
\bar{r}_i &=& \frac{r_{i-1} + r_{i}}{2}, \\
\bar{\theta}_j &=& \frac{\theta_{j-1} + \theta_{j}}{2}.
\end{eqnarray}

Each variable is discretized with respect of its cell as follows:
\begin{itemize}
\item $\varPhi$, $C$ and $P$ are represented by their value at the center of each cell, 
using an all-centered grid:
\begin{eqnarray}
\varPhi^h_{[i,j]} &=& \varPhi(\bar{r}_i, \bar{\theta}_j), \\
C^h_{[i,j]} &=& C(\bar{r}_i, \bar{\theta}_j), \\
P^h_{[i,j]} &=& P(\bar{r}_i, \bar{\theta}_j).
\end{eqnarray}
\item $\bV$ is represented by its values at cell boundaries, using a staggered grid:
\begin{eqnarray}
V_r^h{}_{[i,j]} &=& V_r(r_i, \bar{\theta}_j), \\
V_\theta^h{}_{[i,j]} &=& V_\theta(\bar{r}_i, {\theta}_j).
\end{eqnarray}
\end{itemize}

\subsubsection{Operator Discretization}
A finite-volume method with linear interpolation is used for flux discretization. 
%$\boldsymbol{f}^h$

Define the following discrete central difference operators:
\begin{eqnarray}
\cD_r(f^h){}_{\left[i+\half,j\right]} &=& \frac{f^h_{\left[i+1,j\right]} - f^h_{\left[i,j\right]}}
                       {r_{\left[i+1,j\right]} - r_{\left[i,j\right]}}, \\
\cD_\theta(f^h){}_{\left[i,j+\half\right]} &=& \frac{f^h_{\left[i,j+1\right]} - f^h_{\left[i,j\right]}}
					   {\theta_{\left[i,j+1\right]} - \theta_{\left[i,j\right]}}.
\end{eqnarray}

@@

The total flux of a cell is equal to zero ($\bnabla \cdot \boldsymbol{f} = 0$):
\begin{eqnarray}
\frac{1}{r^2} \cD_r\pars{f^h_r r^2} + 
\frac{1}{r \sin\theta} \cD_\theta\pars{f^h_\theta \sin\theta} =
\frac{\cD_r\pars{f^h_r \cdot r^2 \sin\theta} + 
\cD_\theta\pars{f^h_\theta \cdot r \sin\theta}}{r^2 \sin\theta} = 0. 
\end{eqnarray}

Ion flux ($\bI = - C \bnabla \varPhi$) is discretized on grid cell boundaries:
\begin{eqnarray}
I^h_r &=& -\cI_r(C^h) \cdot \cD_r(\varPhi^h), \\
I^h_\theta &=& -\cI_\theta(C^h) \cdot \frac{\cD_\theta(\varPhi^h)}{r}.
\end{eqnarray}

>>>>

\title{Asymptotic and Numerical Solution for Electro-Kinetic Flow Around an Ion-Selective Particle}

@@

@@

\abstract{A numerical solver for the simulation of an electrokinetic migration of particles in an
electrolyte solution is presented. The electrokinetic transport process is
described by a nonlinear partial differential equation system, 
as well as nonlinear boundary conditions.
The nonlinear system is discretized and solved iteratively using Newton's method.
An asymptotic analytical solution is derived for the validation of the solver. 
The numerical results are compared to the asymptotic solutions, and good correspondence is achieved.}

@@

Electrokinetic theory describes the dynamics of charged particles
in ionic fluids \cite{masliyah2005book,kirby2010book}.
When a particle acquires surface charge, a layer
of ions of opposite charge is attracted to the surface via    
electric forces, creating a double-layer structure around the
particle (see Figure \ref{fig:EDL}). This structure, called
``Debye layer'', electrically screens the surface charge, thus
creating a potential difference between the particle and the outer
layer of the fluid bulk.
In cases where the layer width is much smaller than the particle
size, an analytical asymptotic solution for the Debye layer
dynamics can be found \cite{yariv2010asymptotic}.
\begin{figure}[htbp]
    \begin{center}
        \includegraphics[width=0.2\textwidth]
            {figs/ElectricDoubleLayer.eps}
        \caption{Schematic structure of the double layer:
        (1) particle surface, (2) Debye layer, (3) fluid bulk.
        If the particle surface has positive (blue) surface charge,
        it attracts negative (red) ions from the fluid making the
        Debye layer negatively charged (as opposed to the rest of
        the fluid bulk, which is electrically neutral).
        The zeta potential is defined as the voltage drop across 
        the Debye layer (2).}
        \label{fig:EDL}
    \end{center}
\end{figure}
% \begin{tikzpicture}\end{tikzpicture}

The variables of the electrokinetic problem are the electrostatic
potential $\varphi$, fluid velocity $\bv$ and its pressure $p$, and
ionic concentration $c$.
The boundary conditions are determined by the specific
problem under consideration and defined by the particle's
geometry, chemical characteristics, and the fluid dynamics.
The partial differential equations that describe the system dynamics
under an external electric field, are coupled and nonlinear, and
in general they have no analytic solution. 
Moreover, any numerical solver must handle the scale disparity caused by the
Debye layer width being much smaller than the particle size. 
A closed-form linear asymptotic solution has been developed for
spherical particles and weak electric field in an
axisymmetric setting \cite{yariv2010migration} but it is hard to extend this
analytic solution to more general systems. Once the electric field
becomes stronger, significant nonlinear phenomena are expected.
This interesting regime has not yet been explored.
Recently, it has been conceived that such type of dynamics
can be very useful in transporting and manipulating micro-
and nanoscale objects in many nanotechnology applications,
nano-fluidic devices, packed bed separation, desalination processes and various 
electrophoresis applications \cite{howse2007self,paxton2004catalytic,pumera2010electrochemically,nischang2007,leinweber2004}.
Because the model for these electrokinetic
phenomena has no closed-form analytic solution, a numerical
solver for this system is of importance.

The goal of this work is to develop and implement an iterative numerical
solver for nonlinear electrokinetic problems, and apply it to the study
of systems that have no closed-form solutions.
This solver can be used to gain insight into the chemical and physical behavior 
in far more general regimes than are currently well-understood, 
and also to possibly lead to further analytical developments, based on new scaling behavior
discovered numerically. We may also target interesting physical
phenomena that have been observed experimentally.
One such phenomenon is nonlinear electrokinetic vortex flow, 
which is useful for microfluidic mixing applications \cite{wang2004mix, ben2002vortex},

The paper is structured as follows: the governing equations of the electrokinetic model
are derived in Section \ref{sec:equations}. The algorithm for the discretization and 
the solution of the system is described in Section \ref{sec:algorithm}.
The analytical asymptotic solution for the non-linear system is derived 
and presented in Section \ref{sec:asymp}.
The numerical results and their comparison to an analytical asymptotic solution 
are presented and discussed in Section \ref{sec:results}.

@@

Spatial coordinates $\br$ are normalized by $a^*$ 
(the characteristic size of the ion-exchanging particle).
Positive and negative ionic concentrations are denoted by $c_+$ and $c_-$, respectively, and
are normalized by the ambient ionic concentration $c^*$. The ionic valences are $\pm z$.
The dimensional ionic fluxes, $\bj_\pm$, are given by the Nernst-Plank equation 
(using the Einstein relation for the electric mobility, $\mu_q^* = D^* / k_B^* T^*$):
\begin{eqnarray}
\bj^*_\pm &=& 
-D^* \bnabla c^*_\pm + \bv^* c^*_\pm \mp \frac{z e^* D^*}{k_B^* T^*} c^*_\pm \bnabla \varphi^*,
\end{eqnarray}
where $D^*$ is the ionic diffusivity.

@@

The dimensional Poisson equation can be written as:
\begin{eqnarray}
(c^*_+ - c^*_-) z e^* &=& -\eps^* \Laplacian \varphi^*,
\end{eqnarray}
and after nondimensionalization, is given by:
\begin{eqnarray}
c_+ - c_- &=& -\frac{\eps^* \varphi^*}{z e^* c^* (a^*)^2} \Laplacian \varphi = 
-\delta^2 \Laplacian \varphi,
\\
\delta^2 &=& \frac{\eps^* \varphi^*}{c^* e^* z (a^*)^2} = 
\pars{\frac{\delta^*}{a^*}}^2.
\end{eqnarray}

@@

\subsection{Governing Equations}
The electro-kinetic flow is described by the following nondimensional equations:

@@

Due to conservation of ions, the fluxes are divergence-free:
\begin{eqnarray}
\bnabla \cdot \bj_\pm &=& 0.
\end{eqnarray}
These equation may be re-written by using the following transformations:
\begin{eqnarray}
  c = \frac{c_+ + c_-}{2} &,& q = \frac{c_+ - c_-}{2}.
\end{eqnarray}
Thus, salt and charge fluxes are defined by:
\begin{eqnarray}
  \bj = \frac{\bj_+ + \bj_-}{2} = -\bnabla c - q \bnabla \varphi + \alpha \bv c &,& 
  \bi = \frac{\bj_+ - \bj_-}{2} = -\bnabla q - c \bnabla \varphi + \alpha \bv q,
\end{eqnarray}
and are divergence-free as well:
\begin{eqnarray}
\bnabla \cdot \bj = 0 &,& 
\bnabla \cdot \bi = 0. 
\end{eqnarray}

@@

\subsubsection{Boundary conditions}
We follow the physical setting described in \cite{yariv2010migration}.
We assume that a constant constant electric field of nondimensional magnitude $\beta$,
is applied to the system, in direction denoted by the unit vector $\ui$.
As a result, the ion-exchanger will equilibrate in 
a non-zero steady-state velocity $\cU \ui$, which is
a function of the applied field, and is not known a priori.

We employ the reference frame of the ion exchanger.
Far away from the particle, we have a uniform applied electrostatic field, uniform flow,
and ambient ionic concentrations:
\begin{eqnarray}
\bv \rightarrow -\cU \ui, &
\bnabla \varphi \rightarrow -\beta\ui, &
 c_\pm \rightarrow 1.
\end{eqnarray}

We consider a spherical particle.
The chemical properties of the particle's surface $\mathcal{S}$ (defined at $r=1$ with normal $\bn = \brhat$), determine the boundary conditions on $\mathcal{S}$.
we assume high conductance of the particle (yielding a uniform electric potential),
zero slip, anion impermeability, cation selectivity 
(with fast Butler-Volmer\cite{bard2000book} kinetics
$\bn \cdot \bj_+ = k (1 - c_+/\gamma)$, where $k \rightarrow \infty$):
\begin{eqnarray}
\begin{array}{cccc}
\varphi = \cV, &
\bv = \bzero, &
\bn \cdot \bj_- = 0, &
c_+ = \gamma,
\end{array}
\end{eqnarray}
where $\gamma$ is the ratio between the cationic concentration on the surface and the cationic
concentration in the fluid bulk.

@@

\subsubsection  {Bulk-scale equations}
The Nernst-Planck equations above for the fluid bulk (outside
the boundary layer), using bulk variables denoted by uppercase letters, read:
\begin{eqnarray}
\begin{array}{cccc}
  Q = 0, &
  C = C_+ = C_-, &
\bJ = -\bnabla C + \alpha \bV C, &
\bI = -C \bnabla \varPhi.
\end{array}\end{eqnarray}
Because the flow is incompressible ($\bnabla \cdot V = 0$), 
ion conservation results in the following 
salt and charge conservation equations:
\begin{eqnarray} 
\label{eq:salt}
\bnabla \cdot \bJ = \Laplacian C - \alpha \bV \cdot \bnabla C &=& 0, \\
\label{eq:charge}
\bnabla \cdot \bI = \bnabla \cdot \pars{ C \bnabla \varPhi } &=& 0.
\end{eqnarray}

@@

\subsubsection{Bulk-scale effective boundary conditions}
By integrating the microscale equations across the Debye layer
(see full derivation in \cite{yariv2010asymptotic}), 
the effective boundary conditions at the ion-exchanger surface are:
\begin{eqnarray}
0 &=& \varPhi + \log C, \\
0 &=& \bnhat \cdot \bnabla \pars{\varPhi - \log C}, \\
\bV &=& 
\zeta \cdot \bnabla_\mathcal{S} \varPhi 
+ 2\log\pars{1-\tanh^2\frac{\zeta}{4}} \cdot \bnabla_\mathcal{S} \log C,
\end{eqnarray}
where $\zeta = \cV - \varPhi = \log (C / \gamma)$ 
is the zeta potential (the voltage drop across the Debye layer).

\subsection{Steady state}
The stress tensor is composed of Newtonian and Maxwel stresses:
\begin{eqnarray}
\tT &=& \bnabla \bV + (\bnabla \bV)^\dagger - P \tI
+ \bnabla \varPhi \bnabla \varPhi - \frac{1}{2} (\bnabla \varPhi \cdot \bnabla \varPhi) \tI, \\
\bnabla \cdot \tT &=& \bzero.
\end{eqnarray}
In steady state (constant particle drift velocity $\cU$), the total force acting on the particle 
(computed by the following surface integral) must vanish:
\begin{eqnarray}
\bF = \oint_\mathcal{S} \tT \cdot \bnhat dA = \bzero.
\end{eqnarray}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Algorithm Description} \label{sec:algorithm}
The numerical solver first requires constructing a discrete 
approximation for the nonlinear PDE system,
for a given electric field $\beta$ and a velocity $\cU$. 
The resulting nonlinear discrete system is
solved numerically using Newton's Method, so the total force on the
particle can be computed. 
We seek the steady-state velocity $\cU$, 
which corresponds to zero total force,
\begin{eqnarray}
F(\beta, \cU) = 0,
\end{eqnarray}
which we solve by the standard secant method.

\subsection{Discretization}
\subsubsection{Spherical Coordinates}
Because the system is axisymmetric, we employ
spherical coordinates $(r,\theta,\phi)$.
The scalar gradient and the divergence can be written as:
\begin{eqnarray}
\bnabla f &=& \deriv{f}{r}\brhat + \frac{1}{r}\deriv{f}{\theta}\bthetahat, \\
\bnabla \cdot \bF &=& \frac{1}{r^2}\deriv{}{r}\pars{F_r r^2 } + 
               \frac{1}{r \sin\theta}\deriv{}{\theta}\pars{F_\theta \sin\theta}.
\end{eqnarray}
The scalar Laplacian can be written as:
\begin{eqnarray}
\Laplacian f = \bnabla \cdot (\bnabla f)&=& 
 \frac{1}{r^2}\deriv{}{r}\pars{r^2 \deriv{f}{r}} + 
 \frac{1}{r^2 \sin\theta}\deriv{}{\theta}\pars{\sin\theta \deriv{f}{\theta}}.
\end{eqnarray}
The vector gradient can be written as:
\begin{eqnarray}
\bnabla \bF &=& \deriv{F_r}{r} \brhat \brhat + \deriv{F_\theta}{r} \brhat \bthetahat + 
\frac{1}{r}\pars{\deriv{F_r}{\theta} - F_\theta} \bthetahat \brhat + 
\frac{1}{r}\pars{\deriv{F_\theta}{\theta} + F_r} \bthetahat \bthetahat.
\end{eqnarray}
The vector Laplacian can be written as:
\begin{eqnarray}
\bLaplacian \bF &=& 
\left(\Laplacian F_r - \frac{2F_r}{r^2} - 
\frac{2}{r^2 \sin\theta} \deriv{\left(F_\theta \sin\theta \right)}{\theta}\right)\brhat
+ \left(\Laplacian F_\theta - \frac{F_\theta}{r^2 \sin^2\theta} + 
\frac{2}{r^2}\deriv{F_r}{\theta}\right) \bthetahat.
\end{eqnarray}

Note that in the spherical coordinate system,
the vector Laplacian operator cannot be computed in a decoupled manner as in the Cartesian system,
since the unit vectors $\brhat, \bthetahat, \bphihat$ are functions of the coordinates
in a curvilinear coordinate system.

\subsubsection{Computational Grid}
A regular grid of size $n_r \times n_\theta$ is defined, with:
\begin{eqnarray}
(r_i,\theta_j) \in [1, \infty) \times [0,\pi], & 
r_i = (1+\Delta_r)^i, &
 \theta_j = \Delta_\theta \cdot j,
\end{eqnarray}
where a logarithmic grid spacing has been chosen for $r$ and a uniform grid for $\theta$.
Note that $r_0 = 1$ and $R_{max} = (1+\Delta_r)^{n_r}$ should be much 
larger than $r_0$. Thus:
\begin{eqnarray}
\Delta_r = \pars{R_{max}} ^ \frac{1}{n_r} - 1 ,& \Delta_\theta = \frac{\pi}{n_\theta}.
\end{eqnarray}

This grid induces a disjoint subdivision of the domain 
$\Omega = [1, R_{max}] \times [0,\pi]$ into cells:
\begin{eqnarray}
\bigcup_{ij}\Omega_{ij} &=& \Omega.
\end{eqnarray}
A specific cell $\Omega_{ij}$ and its center $(\bar{r}_i, \bar{\theta}_j)$ are defined by:
\begin{eqnarray}
\Omega_{ij} = [r_{i-1}, r_{i}] \times [\theta_{j-1}, \theta_{j}], &
\bar{r}_i = (r_{i-1} + r_{i})/2, &
\bar{\theta}_j = (\theta_{j-1} + \theta_{j})/2.
\end{eqnarray}

Each discrete variable is located with respect of its cell as follows:
$\varPhi$, $C$ and $P$ are represented by their value at the center of each cell, 
using an all-centered grid:
\begin{eqnarray}
\varPhi^h_{[i,j]} = \varPhi(\bar{r}_i, \bar{\theta}_j), &
C^h_{[i,j]} = C(\bar{r}_i, \bar{\theta}_j), &
P^h_{[i,j]} = P(\bar{r}_i, \bar{\theta}_j).
\end{eqnarray}
$\bV$ is represented by its values at cell boundaries, using a staggered grid:
\begin{eqnarray}
V_r^h{}_{[i,j]} = V_r(r_i, \bar{\theta}_j), &
V_\theta^h{}_{[i,j]} = V_\theta(\bar{r}_i, {\theta}_j).
\end{eqnarray}

\subsubsection{Operator Discretization}
A finite-volume method with linear interpolation is used for flux discretization.

Define the following discrete central difference operators:
\begin{eqnarray}
\cD_r(f^h){}_{\left[i+\half,j\right]} = \frac{f^h_{\left[i+1,j\right]} - f^h_{\left[i,j\right]}}
                       {r_{\left[i+1,j\right]} - r_{\left[i,j\right]}}, &
\cD_\theta(f^h){}_{\left[i,j+\half\right]} = \frac{f^h_{\left[i,j+1\right]} - f^h_{\left[i,j\right]}}
					   {\theta_{\left[i,j+1\right]} - \theta_{\left[i,j\right]}}.
\end{eqnarray}

@@

If the total flux of a cell is equal to zero ($\bnabla \cdot \boldsymbol{f} = 0$), then we have
\begin{eqnarray}
\frac{1}{r^2} \cD_r\pars{f^h_r r^2} + 
\frac{1}{r \sin\theta} \cD_\theta\pars{f^h_\theta \sin\theta} =
\frac{\cD_r\pars{f^h_r \cdot r^2 \sin\theta} + 
\cD_\theta\pars{f^h_\theta \cdot r \sin\theta}}{r^2 \sin\theta} = 0. 
\end{eqnarray}

Ion flux ($\bI = - C \bnabla \varPhi$) is discretized on grid cell boundaries:
\begin{eqnarray}
I^h_r = -\cI_r(C^h) \cdot \cD_r(\varPhi^h), &
I^h_\theta = -\cI_\theta(C^h) \cdot \frac{\cD_\theta(\varPhi^h)}{r}.
\end{eqnarray}

Salt flux ($\bJ = -\bnabla C + \alpha \bV C$) is discretized on grid cell boundaries 
(using upwind scheme $\mathcal{U}$ for numerical stability at large cell Peclet number):
\begin{eqnarray}
J^h_r = -\cD_r(C^h) + \alpha V^h_r \cdot \mathcal{U}^{\bV^h}_r (C^h), &
J^h_\theta = -\frac{\cD_\theta(C^h)}{r} + \alpha V^h_\theta \cdot \mathcal{U}^{\bV^h}_\theta (C^h), 
\\
 \mathcal{U}^{\bV^h}_r(C^h)[i,j] = C^h\left[i-\frac{\sign(V^h_r)}{2}, j\right], &
 \mathcal{U}^{\bV^h}_\theta(C^h)[i,j] = C^h\left[i, j-\frac{\sign(V^h_\theta)}{2}\right]. 
\end{eqnarray}

Mass flux $\bV$ is discretized on grid cells boundaries, 
using the staggered velocity grid.
